\bf A class of exponential family measurement error models for singe-cell CRISPR screens

### 2022-01-04 text/GLM-EIV/jasa.tex
fix reproducibility

\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{setspace}
\usepackage{amsfonts}
\usepackage{comment}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{tabularx}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{array}
\usepackage{titlesec} 
\usepackage{xcolor}
\usepackage{float}
\usepackage[toc,page]{appendix}
\usepackage{algpseudocode}
\usepackage{afterpage}
\usepackage[
backend=biber,
style=numeric-comp,
maxcitenames=1,
maxbibnames=3,
backref=true,
doi=false,
isbn=false,
url=false,
eprint=false
]{biblatex}

% MY COMMANDS
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\addbibresource{/Users/timbarry/optionFiles/glmeiv.bib}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\newcommand{\PreserveBackslash}[1]{\let\temp=\\#1\let\\=\temp}
\newcolumntype{C}[1]{>{\PreserveBackslash\centering}p{#1}}
\newcommand{\indep}{\raisebox{0.05em}{\rotatebox[origin=c]{90}{$\models$}}}

\makeatletter
\newcommand{\multiline}[1]{%
	\begin{tabularx}{\dimexpr\linewidth-\ALG@thistlm}[t]{@{}X@{}}
		#1
	\end{tabularx}
}
\makeatother

% JOURNAL SETTINGS
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{1}
\addtolength{\oddsidemargin}{-.5in}
\addtolength{\evensidemargin}{-1in}
\addtolength{\textwidth}{1in}
\addtolength{\textheight}{1.7in}
\addtolength{\topmargin}{-1in}
\date{\vspace{-5ex}}

\begin{document}
\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}
\if1\blind
{
  \title{\bf A new class of measurement error models for CRISPR genome engineering and single-cell RNA sequencing}
  \author{Timothy Barry$^1$, Eugene Katsevich$^2$, Kathyrn Roeder$^1$ \\
    $^1$Department of Statistics and Data Science, Carnegie Mellon University \\
    $^2$Department of Statistics and Data Science, Wharton School}
  \maketitle
} \fi

\if0\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE \bf A new class of measurement error models for CRISPR genome engineering and single-cell RNA sequencing}
\end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
CRISPR genome engineering and single-cell RNA sequencing have transformed biological discovery. Single-cell CRISPR screens unite these two technologies, linking genetic perturbations in individual cells to changes in gene expression and illuminating regulatory networks underlying diseases. Despite their promise, single-cell CRISPR screens present substantial statistical challenges. We demonstrate through theoretical and real data analyses that a standard method for estimation and inference in single-cell CRISPR screens --``thresholded regression'' -- exhibits attenuation bias and a bias-variance tradeoff as a function of an intrinsic and challenging-to-select tuning parameter. To overcome these difficulties, we introduce GLM-EIV (``GLM-based errors-in-variables''), a new method for single-cell CRISPR screen analysis. GLM-EIV extends the classical errors-in-variables model to responses and noisy predictors that are (i) exponential family-distributed and (ii) potentially impacted by the same set of confounding variables. We develop a computational infrastructure to deploy GLM-EIV across tens or hundreds of nodes on clouds (e.g., Microsoft Azure) and high-performance clusters. Leveraging this infrastructure, we apply GLM-EIV to analyze two recent, large-scale, single-cell CRISPR screen datasets, demonstrating improved performance in challenging problem settings.
\end{abstract}

\noindent
{\it Keywords:}  exponential family, errors-in-variables, GLM, CRISPR screen, differential expression
\spacingset{1.9}
\section{Introduction}
CRISPR is a genome engineering tool that has enabled scientists to precisely edit human and nonhuman genomes, opening the door to new medical therapies \cite{Rothgangl2021,Musunuru2021} and transforming basic biology research \cite{Przybyla2021}. Recently, scientists have paired CRISPR genome engineering with single-cell sequencing \cite{Dixit2016,Datlinger2017}. The resulting assays, known as a ``single-cell CRISPR screens,'' link genetic perturbations in individual cells to changes in gene expression, illuminating regulatory networks underlying human diseases and other traits \cite{Morris2021a}.

Despite their promise, single-cell CRISPR screens present substantial statistical challenges. A major difficulty is that CRISPR perturbations are assigned stochastically to cells and cannot be observed directly. As a consequence, one cannot know with certainty which cells were perturbed. Instead, one must leverage an indirect, noisy proxy of perturbation presence or absence -- namely, transcribed guide RNA counts -- to ``guess'' which cells were perturbed. Using these imputed perturbation assignments, one can attempt to estimate the effect of the perturbation on gene expression. The standard approach, which we call ``thresholded regression'' or the ``thresholding method,'' is to assign perturbation identities to cells by simply thresholding the guide RNA counts.

We study estimation and inference in single-cell CRISPR screens from a statistical perspective, formulating the data generating mechanism using a new class of measurement error models. We assume that the response variable $y$ is a GLM of an underlying predictor variable $x^*$ and vector of confounders $z$. We do not observe $x^*$ directly; rather, we observe a noisy version $x$ of $x^*$ that itself is a GLM of $x^*$ and $z$. The goal of the analysis is to estimate the effect of $x^*$ on $y$ using the observed data $(x, y, z)$ only. In the context of the biological application, $x^*$, $x$, $y$, and $z$ are CRISPR perturbations, guide RNA counts, gene expressions, and technical confounders, respectively.

Our work makes two main contributions. First, we conduct a detailed study of the thresholding method. Notably, we demonstrate on real data that the thresholding method exhibits attenuation bias and a bias-variance tradeoff as a function of the selected threshold, and we recover these phenomena in precise mathematical terms in a simplified Gaussian setting. Second, we introduce a new method, GLM-EIV (``GLM-based errors-in-variables''), for single-cell CRISPR screen analysis. GLM-EIV extends the classical errors-in-variables model to response distributions and sources of measurement error that are (i) exponential family-distributed and (ii) potentially impacted by the same set of confounding variables. GLM-EIV thereby implicitly estimates the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding or another heuristic. % Theoretical analyses and simulation studies indicate that GLM-EIV outperforms the thresholding method in large regions of the parameter space.
We implement several statistical accelerations (that possibly are of independent utility) to bring the cost of GLM-EIV down to within about an order of magnitude of the thresholding method.

@@

\section{Background and analysis challenges}

\subsection{Related work}
Motivated by the challenges of single-cell data, several authors recently have extended statistical models that (implicitly or explicitly) assume Gaussianity and homoscedasticity to a broader class of exponential family distributions. For example, Lin et al.\ \cite{Lin2021} and Townes et al.\ \cite{Townes2019}  (separately) developed eSVD and GLM-PCA, generalizations of SVD and PCA, respectively, to exponential family response distributions. Unlike their vanilla counterparts, eSVD and GLM-PCA can model gene expression counts directly, thereby capturing the relationship between the mean and variance of a gene's expression level \cite{Lause2021} and improving performance on dimension reduction tasks. We see our work as a continuation of this effort to ``port'' common statistical methods and models to single-cell count data. Our focus, however, is on regression rather than dimension reduction. We extend the classical errors-in-variables model in several directions, most importantly generalizing to exponential family-distributed response distributions and sources of measurement error.

@@

The genomics literature has produced several applied methods for single-cell CRISPR screen analysis. Barry et al.\ developed SCEPTRE \cite{Barry2021}, a custom implementation of the conditional randomization test \cite{Candes2018, Liu2021} tailored to single-cell CRISPR screen data. SCEPTRE tests whether a given perturbation is associated with the change in expression of a given gene, adjusting for sources of confounding and ensuring robustness to expression model misspecification. In this work we tackle a set of analysis challenges that are complimentary to those addressed by SCEPTRE. Most importantly, we account for the fact that the perturbation is measured with noise, an issue that SCEPTRE circumvents via thresholding. Additionally, we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, an objective that is challenging to attain within the nonparametric hypothesis testing framework of SCEPTRE.

%Other applied methods for single-cell CRISPR screen analysis include MIMOSCA \cite{Dixit2016} and scMAGeCK \cite{Yang2019}. % These methods, like SCEPTRE, focus on hypothesis testing (rather than estimation), but unlike SCEPTRE, they ignore the countedness of the data and are unable to handle confounders.

\subsection{Assay overview}

There are several broad classes of single-cell CRISPR screen assays, each suited to answer a different set of biological questions \cite{Gasperini2019,Datlinger2021,Mimitou2019}. In this work we focus on high-multiplicity of infection (MOI) single-cell CRISPR screens. We expect the ideas that we develop for this assay to apply (with some effort) to other classes of single-cell CRISPR screens as well. In this section we motivate high MOI single-cell screens, overview the experimental protocol, and present relevant analysis challenges.

The human genome consists of genes, enhancers (segments of DNA that regulate the expression of one or more genes), and other genomic elements (that are not of importance to the current work). Genome-wide association studies (GWAS) have revealed that the majority ($>90\%$) of variants associated with diseases lie outside genes and (very likely) inside enhancers \cite{Gallagher2018}. These noncoding variants are thought to contribute to disease by modulating the expression of one or more disease-relevant genes. Scientists do not know the gene (or genes) through which most noncoding variants exert their effect, limiting the interpretability of GWAS results. A central open challenge in genetics, therefore, is to link enhancers that harbor GWAS variants to the genes that they target at genome-wide scale \cite{Gasperini2020}.

High MOI single-cell CRISPR screens are the most promising biotechnology for solving this challenge. High MOI single-cell CRISPR screens combine CRISPR interference (CRISPRi) -- a version of CRISPR that represses a targeted region of the genome -- with single-cell sequencing. The experimental protocol is as follows. First, the scientist develops a library of several hundred to several thousand CRISPRi perturbations, each designed to target a candidate enhancer for repression. The scientist then cultures tens or hundreds of thousands of cells and delivers the CRISPRi perturbations to these cells. The perturbations assort into the cells randomly, with each cell receiving on average 10-40 distinct perturbations. Conversely, a given perturbation enters about 0.1-2\% of cells.

@@

\subsection{Analysis challenges}
High MOI single-cell CRISPR screens present several statistical challenges, four of which we highlight here. Throughout, we consider a single perturbation-gene pair. First, the ``treatment'' variable -- i.e., the presence or absence of a perturbation -- cannot be directly observed. Instead, perturbed cells transcribe molecules called  \textit{guide RNAs} (or \textit{gRNAs}) that serve as indirect proxies of perturbation presence. We must leverage these gRNAs to impute (explicitly or implicitly) perturbation assignments onto the cells (Figure \ref{analysis_challenges}b). Second, ``technical factors'' -- sources of variation that are experimental rather than biological in origin -- impact the measurement of both gene and gRNA expressions and therefore act as confounders (Figure \ref{analysis_challenges}b). Third, the gene and gRNA data are sparse, discrete counts. Consequently, classical statistical approaches that assume Gaussianity or homoscedasticity are inapplicable. Finally, and most subtly, sequenced gRNAs sometimes map to cells that have not received a perturbation. This phenomenon, which we ``background contamination,'' results from errors in the sequencing and alignment processes \cite{Replogle2020}. The marginal distribution of the gRNA counts is best conceptualized as a mixture model (Figure \ref{analysis_challenges}c; Gaussian distributions used for illustration purposes only). Unperturbed and perturbed cells both exhibit nonzero gRNA count distributions, but this distribution is shifted upward for perturbed cells. Figure \ref{analysis_challenges}d shows example data on four (of possibly tens or hundreds of thousands of) cells. The analysis objective is to leverage the gene expressions and gRNA counts to estimate the effect of the (latent) perturbation on gene expression, accounting for the technical factors.

In this work we analyze two large-scale, high MOI, single-cell CRISPR screen datasets published by Gasperini et al.\ \cite{Gasperini2019} and Xie et al.\ \cite{Xie2019a}. Gasperini (resp., Xie) targeted approximately 6,000 (resp., 500) candidate enhancers in a population of approximately 200,000 (resp., 100,000) cells. Gasperini additionally designed several hundred positive control, gene-targeting perturbations and 50 non-targeting, negative control perturbations to assess method sensitivity and specificity.

\section{Thresholding method}
We study thresholding from empirical and theoretical perspectives, highlighting several inherent limitations of the approach. Gasperini and Xie imputed perturbation assignments onto the cells via thresholding, but they carried out the subsequent differential expression analysis in different ways: Gasperini fitted negative binomial regression models to the data, whereas Xie applied nonparametric tests of independence. We study Gasperini's variant of the thresholding method, as it relates more closely to GLM-EIV and (in our view) is the more natural method.

Let $n \in \mathbb{N}$ be the number of cells assayed in the experiment. Consider a single perturbation and a single gene. For cell $i \in \{1, \dots, n\}$, let $m_i \in \mathbb{N}$ be the number of gene transcripts sequenced; let $g_i \in \mathbb{N}$ be the number of gRNA transcripts sequenced; let $d^m_i \in \mathbb{N}$ be the number of gene transcripts sequenced across \textit{all} genes (i.e., the library size or sequencing depth); and finally, let $z_i \in \mathbb{R}^{d-1}$ be the cell-specific technical factors (e.g., sequencing batch, percent mitochondrial reads, etc.) The letters ``m,'' ``g'', and ``d'' stand for ``mRNA,'' ``gRNA,'' and ``depth,'' respectively. The thresholding method is defined as follows:
\begin{itemize}
	\item[1.] For a given threshold $c \in \mathbb{N}$, let the imputed perturbation assignment $\hat{p}_i \in \{0, 1\}$ be $$ \hat{p}_i = 0 \textrm{ if } g_i < c; \quad \hat{p}_i = 1 \textrm{ if } g_i \geq c.$$
	%$$\begin{cases} \hat{p}_i = 0 \textrm{ if } g_i < c, \\ \hat{p}_i = 1 \textrm{ if } g_i \geq c. \end{cases}$$
	\item[2.] Assume that $m_i$ is related to $\hat{p}_i, d^m_i,$ and $z_i$ through the following GLM:
	\begin{equation}\label{thresh_glm}
	m_i | (\hat{p}_i, z_i, d^m_i) \sim \textrm{NB}_{\theta^m}(\mu_i); \quad
	\log(\mu_i) = \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i + \log\left(d_i^m\right),
	\end{equation}
	where (i) $NB_{\theta^m}(\mu_i)$ is a negative binomial distribution with mean $\mu_i$ and known size parameter $\theta^m$; (ii) $\beta^m_0 \in \mathbb{R}, \beta^m_1 \in \mathbb{R},$ and $\gamma_m \in  \mathbb{R}^{d-1}$ are unknown parameters; and (iii) $\log(d_i^m)$ is an offset term.

	\item[3.]  Fit a GLM to obtain an estimate and confidence interval for the target of inference $\beta^m_1$.
\end{itemize}
The sequencing depth $d^m_i$ is included as an offset term in (\ref{thresh_glm}) so that $\beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i$ can be interpreted as a relative expression. Exponentiating both sides of (\ref{thresh_glm}) yields
$$\mu_i = \exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right) d_i^m.$$ We see that $\exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right)$ is the \textit{fraction} of all transcripts sequenced in the cell produced by the gene under consideration.

Fold change in this context is defined as the ratio of mean gene expression in perturbed cells to mean gene expression in unperturbed cells. The target of inference $\beta^m_1$ is the \textit{log} fold change in expression in response to the perturbation, controlling for the technical factors. Therefore, $\exp(\beta^m_1) = 1$ (eq., $\beta^m_1 = 0$) indicates no change in expression, whereas $\exp(\beta^m_1) > 1$ (eq., $\beta^m_1 > 0$) and $\exp(\beta^m_1) < 1$ (eq., $\beta^m_1 < 0$) indicate an increase and decrease in expression, respectively.

\subsection{Empirical challenges of the thresholding method}\label{sec:thresholding_empirical}

>>>>

\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{setspace}
\usepackage{amsfonts}
\usepackage{comment}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{tabularx}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{array}
\usepackage{titlesec} 
\usepackage{xcolor}
\usepackage{float}
\usepackage[toc,page]{appendix}
\usepackage{algpseudocode}
\usepackage{afterpage}
\usepackage[
backend=biber,
style=authoryear,
maxcitenames=1,
maxbibnames=3,
uniquelist=false,
backref=true,
doi=false,
isbn=false,
url=false,
eprint=false
]{biblatex}

% MY COMMANDS
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\addbibresource{glmeiv.bib}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\newcommand{\PreserveBackslash}[1]{\let\temp=\\#1\let\\=\temp}
\newcolumntype{C}[1]{>{\PreserveBackslash\centering}p{#1}}
\newcommand{\indep}{\raisebox{0.05em}{\rotatebox[origin=c]{90}{$\models$}}}
\allowdisplaybreaks
\makeatletter
\newcommand{\multiline}[1]{%
	\begin{tabularx}{\dimexpr\linewidth-\ALG@thistlm}[t]{@{}X@{}}
		#1
	\end{tabularx}
}
\makeatother

% JOURNAL SETTINGS
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{1}
\addtolength{\oddsidemargin}{-.5in}
\addtolength{\evensidemargin}{-1in}
\addtolength{\textwidth}{1in}
\addtolength{\textheight}{1.7in}
\addtolength{\topmargin}{-1in}

\begin{document}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}
\if1\blind
{
	%  \title{\bf FDR control following filtering for hierarchically structured hypotheses}
	\title{\bf A class of exponential family measurement error models for singe-cell CRISPR screens}
	%	Eugene Katsevich, Chiara Sabatti, Marina Bogomolov}
	\author{Timothy Barry\\
		Department of Statistics and Data Science,\\Carnegie Mellon University\\
		and\\
		Eugene Katsevich\\
		Department of Statistics and Data Science,\\Wharton School, University of Pennsylvania\\
		and\\
		Kathryn Roeder\thanks{ FUNDING INFORMATION GOES HERE.}\hspace{.2cm}\\
		Department of Statistics and Data Science,\\Computational Biology Department,\\ 
		Carnegie Mellon University}
	\maketitle
} \fi

\if0\blind
{
	\bigskip
	\bigskip
	\bigskip
	\begin{center}
		{\LARGE\bf Title}
	\end{center}
	\medskip
} \fi

\bigskip
\begin{abstract}
CRISPR genome engineering and single-cell RNA sequencing have transformed biological discovery. Single-cell CRISPR screens unite these two technologies, linking genetic perturbations in individual cells to changes in gene expression and illuminating regulatory networks underlying diseases. Despite their promise, single-cell CRISPR screens present substantial statistical challenges. We demonstrate through theoretical and real data analyses that a standard method for estimation and inference in single-cell CRISPR screens --``thresholded regression'' -- exhibits attenuation bias and a bias-variance tradeoff as a function of an intrinsic, challenging-to-select tuning parameter. To overcome these difficulties, we introduce GLM-EIV (``GLM-based errors-in-variables''), a new method for single-cell CRISPR screen analysis. GLM-EIV extends the classical errors-in-variables model to responses and noisy predictors that are exponential family-distributed and potentially impacted by the same set of confounding variables. We develop a computational infrastructure to deploy GLM-EIV across tens or hundreds of nodes on clouds (e.g., Microsoft Azure) and high-performance clusters. Leveraging this infrastructure, we apply GLM-EIV to analyze two recent, large-scale, single-cell CRISPR screen datasets, demonstrating improved performance in challenging problem settings.
\end{abstract}

\noindent%
{\it Keywords:} genome engineering, cloud computing, exponential family, GLM, latent variable model, scRNA-seq

\spacingset{1.9}

\section{Introduction}

CRISPR is a genome engineering tool that has enabled scientists to precisely edit human and nonhuman genomes, opening the door to new medical therapies \parencite{Rothgangl2021,Musunuru2021} and transforming biological discovery \parencite{Przybyla2021}. Recently, scientists have paired CRISPR genome engineering with single-cell RNA sequencing \parencite{Dixit2016,Datlinger2017}. The resulting assays, known as a ``single-cell CRISPR screens,'' link genetic perturbations in individual cells to changes in gene expression, illuminating regulatory networks underlying human diseases and other traits \parencite{Morris2021a}.

Despite their promise, single-cell CRISPR screens present substantial statistical challenges. A major difficulty is that the ``treatment'' -- i.e., the presence or absence of a CRISPR perturbation -- is assigned randomly to cells and is not directly observable. As a consequence, one cannot know with certainty which cells were perturbed. Instead, one must leverage an indirect, quantitative proxy of perturbation presence or absence to ``guess'' which cells received a perturbation. This indirect proxy takes the form of a so-called guide RNA count, with higher counts indicating that a cell is more likely to have been perturbed. The standard approach to single-cell CRISPR screen analysis is to impute perturbation assignments onto the cells by simply thresholding the guide RNA counts; using these imputations, one can attempt to estimate the effect of the perturbation on gene expression. We call this standard approach ``thresholded regression'' or the ``thresholding method.''

We study estimation and inference in single-cell CRISPR screens from a statistical perspective, formulating the data generating mechanism using a new class of measurement error models. We assume that the response variable $y$ is a GLM of an underlying predictor variable $x^*$ and vector of confounders $z$. We do not observe $x^*$ directly; rather, we observe a noisy version $x$ of $x^*$ that itself is a GLM of $x^*$ and the same set of confounders $z$. The goal of the analysis is to estimate the effect of $x^*$ on $y$ using the observed data $(x, y, z)$ only. In the context of the biological application, $x^*$, $x$, $y$, and $z$ are CRISPR perturbations, guide RNA counts, gene expressions, and technical confounders, respectively.

Our work makes two main contributions. First, we conduct a detailed study of the thresholding method. Notably, we demonstrate on real data that the thresholding method exhibits attenuation bias and a bias-variance tradeoff as a function of the selected threshold, and we recover these phenomena in precise mathematical terms in a simplified Gaussian setting. Second, we introduce a new method, GLM-EIV (``GLM-based errors-in-variables''), for single-cell CRISPR screen analysis. GLM-EIV extends the classical errors-in-variables model \parencite{carroll} to responses and noisy predictors that are exponential family-distributed and potentially impacted by the same set of confounding variables. GLM-EIV thereby implicitly estimates the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding. We implement several statistical accelerations (that possibly are of independent utility) to bring the cost of GLM-EIV down to within about an order of magnitude of the thresholding method.

@@

\section{Background and analysis challenges}
\iffalse
\textcolor{red}{=====================================}
\textcolor{blue}{I don't think this section works as is.  I think the whole section should be relocated and revised, but let me say what you might do, if you keep it here. First, as much as I like the eSVD paragraph, I would argue that we aren't the only ones using negative binomial models for CRISPR screen analysis. For one, SCEPTRE did! So quite honestly I don't think this paragraph is appropriate (but see note below). With a heading like "Related work" people might expect to know what is being done in practice.  Your SCEPTRE paragraph is the place to start, but I think we need to expand the lead sentence.  Cite them all and then give a bit about at least one method available in the literature (very very little detail). I'd pick the Monacle based concept and just go with the idea that it is a negative binomial model that uses thresholding and that it sometimes doesn't work and results in a lot of false signals.  Next you can the remainder of your SCEPTRE paragraph, but again I think it makes sense to say something about the basic model from which the CRT is built.  In other words you want to convey that his method still uses thresholds, but solves the false positive problem.  Now you are ready for your final few sentences about how you'll move beyond this. NOTE: Regarding your eSVD paragraph, you could use something like this paragraph in the Discussion, where you could make the point that it is a good thing when models are built on the generating distribution of the data (rather than forcing a normal by transformation).  Then you could say this paper is like the others you cite in doing this.  I make a try below, see the blue paragraph.\\
Here is the alternative idea.  Since my suggestion causes this section to be only one paragraph, you might consider having Assay Overview first and then putting this material at the lead of analysis challenges without a major heading.  If you moved it to there you could do a better job of citing other works for analysis of CRISPR screens.  I am a fan of giving credit to everyone, but doing so with terse sentences that say something like "Several methods have appeared in the literature, each with some shortcomings (ref, ref, ref).".  Anyway, this section could lead people to believe we have very big egos, basically citing ourselves and almost no one else.}
\textcolor{blue}{THIS IS HOW YOU MIGHT REWRITE THE SCEPTRE PARAGRAPH.The genomics literature has produced a few applied methods for single-cell CRISPR screen analysis, each with some shortcomings (cite Gasp, Xie, etc + Barry). The idea is to test whether a given perturbation is associated with the change in expression of a given gene, while ideally adjusting for sources of confounding and ensuring robustness to expression model misspecification. Most relevant is SCEPTRE, a custom implementation of the conditional randomization test \parencite{Candes2018, Liu2021} tailored to single-cell CRISPR screen data \parencite{Barry2021}.  SCEPTRE successfully handles the challenges of poor calibration and low power shared by other methods.  In this work we tackle a set of analysis challenges that are complimentary to those addressed by SCEPTRE. Most importantly,  we account for the fact that the perturbation is measured with noise, an issue that available methods circumvent via thresholding. Additionally, we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, an objective that is challenging to attain within the nonparametric hypothesis testing framework of SCEPTRE.}
\fi

\subsection{Related work}
Motivated by the challenges of single-cell data, several authors recently have extended statistical models that (implicitly or explicitly) assume Gaussianity and homoscedasticity to a broader class of exponential family distributions. For example, Lin et al.\ \parencite{Lin2021} and Townes et al.\ \parencite{Townes2019}  (separately) developed eSVD and GLM-PCA, generalizations of SVD and PCA, respectively, to exponential family response distributions. Unlike their vanilla counterparts, eSVD and GLM-PCA can model gene expression counts directly, thereby capturing the relationship between the mean and variance of a gene's expression level \parencite{Lause2021} and improving performance on dimension reduction tasks. We see our work as a continuation of this effort to ``port'' common statistical methods and models to single-cell count data. Our focus, however, is on measurement error regression models rather than dimension reduction.

@@

The genomics literature has produced a few applied methods for single-cell CRISPR screen analysis. Most relevantly, Barry et al.\ developed SCEPTRE, a custom implementation of the conditional randomization test \parencite{Candes2018, Liu2021} tailored to single-cell CRISPR screen data \parencite{Barry2021}. SCEPTRE tests whether a given perturbation is associated with the change in expression of a given gene, adjusting for sources of confounding and ensuring robustness to expression model misspecification. In this work we tackle a set of analysis challenges that are complimentary to those addressed by SCEPTRE. Most importantly, we account for the fact that the perturbation is measured with noise, an issue that SCEPTRE circumvents via thresholding. Additionally, we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, an objective that is challenging to attain within the nonparametric hypothesis testing framework of SCEPTRE.

\subsection{Assay background}
There are several broad classes of single-cell CRISPR screen assays, each suited to answer a different set of biological questions \parencite{Gasperini2019,Datlinger2021,Mimitou2019}. In this work we focus on high-multiplicity of infection (MOI) single-cell CRISPR screens, which we motivate and describe here. The human genome consists of genes, enhancers (segments of DNA that regulate the expression of one or more genes), and other genomic elements. Genome-wide association studies (GWAS) have revealed that the majority ($>90\%$) of variants associated with diseases lie outside genes and (very likely) inside enhancers \parencite{Gallagher2018}. These noncoding variants are thought to contribute to disease by modulating the expression of one or more disease-relevant genes. Scientists do not know the gene (or genes) through which most noncoding variants exert their effect, limiting the interpretability of GWAS results. A central open challenge in genetics, therefore, is to link enhancers that harbor GWAS variants to the genes that they target at genome-wide scale \parencite{Gasperini2020}.

High MOI single-cell CRISPR screens are the most promising biotechnology for solving this challenge. High MOI single-cell CRISPR screens combine CRISPR interference (CRISPRi) -- a version of CRISPR that represses a targeted region of the genome -- with single-cell sequencing. The experimental protocol is as follows. First, the scientist develops a library of several hundred to several thousand CRISPRi perturbations, each designed to target a candidate enhancer for repression. The scientist then cultures tens or hundreds of thousands of cells and delivers the CRISPRi perturbations to these cells. The perturbations assort into the cells randomly, with each cell receiving on average 10-40 distinct perturbations. Conversely, a given perturbation enters about 0.1-2\% of cells (this work).

@@

\begin{figure}[h!]
	\centering
	\includegraphics[width=1\linewidth]{figures/analysis_challenges.pdf}
	\caption{\textbf{Experimental design and analysis challenges}: \textbf{a,} Experimental design. For a given perturbation (e.g., the perturbation indicated in blue), we partition the cells into two groups: perturbed and unperturbed. Next, for a given gene, we conduct a differential expression analysis across the two groups, yielding an estimate of the impact of the given perturbation on the given gene. \textbf{b,} DAG representing all variables in the system. The perturbation (latent) impacts both gene expression and gRNA expression; technical factors act as confounders, also impacting gene and gRNA expression. The target of estimation is the effect of the perturbation on gene expression. \textbf{c,} Schematic illustrating the ``background read'' phenomenon. Due to errors in the sequencing and alignment processes, unperturbed cells exhibit a nonzero gRNA count distribution (bottom). The target of estimation is the change in mean gene expression in response to the perturbation (top). \textbf{d}, Example data on four cells for a given perturbation-gene pair. Note that (i) the perturbation is unobserved, and (ii) the gene and gRNA data are discrete counts.}
	\label{analysis_challenges}
\end{figure}

High MOI single-cell CRISPR screens present several statistical challenges, four of which we highlight here. Throughout, we consider a single perturbation-gene pair. First, the ``treatment'' variable -- i.e., the presence or absence of a perturbation -- cannot be directly observed. Instead, perturbed cells transcribe molecules called  \textit{guide RNAs} (or \textit{gRNAs}) that serve as indirect proxies of perturbation presence. We must leverage these gRNAs to impute (explicitly or implicitly) perturbation assignments onto the cells (Figure \ref{analysis_challenges}b). Second, ``technical factors'' -- sources of variation that are experimental rather than biological in origin -- impact the measurement of both gene and gRNA expressions and therefore act as confounders (Figure \ref{analysis_challenges}b). Third, the gene and gRNA data are sparse, discrete counts. Consequently, classical statistical approaches that assume Gaussianity or homoscedasticity are inapplicable. Finally, and most subtly, sequenced gRNAs sometimes map to cells that have not received a perturbation. This phenomenon, which we ``background contamination,'' results from errors in the sequencing and alignment processes \parencite{Replogle2020}. The marginal distribution of the gRNA counts is best conceptualized as a mixture model (Figure \ref{analysis_challenges}c; Gaussian distributions used for illustration purposes only). Unperturbed and perturbed cells both exhibit nonzero gRNA count distributions, but this distribution is shifted upward for perturbed cells. Figure \ref{analysis_challenges}d shows example data on four (of possibly tens or hundreds of thousands of) cells. The analysis objective is to leverage the gene expressions and gRNA counts to estimate the effect of the (latent) perturbation on gene expression, accounting for the technical factors.

In this work we analyze two large-scale, high MOI, single-cell CRISPR screen datasets published by \parencite{Gasperini2019} and \parencite{Xie2019a}. Gasperini (resp., Xie) targeted approximately 6,000 (resp., 500) candidate enhancers in a population of approximately 200,000 (resp., 100,000) cells. Gasperini additionally designed several hundred positive control, gene-targeting perturbations and 50 non-targeting, negative control perturbations to assess method sensitivity and specificity.

\section{Thresholding method}
We study thresholding from empirical and theoretical perspectives, highlighting several limitations of the approach. Gasperini and Xie imputed perturbation assignments onto the cells via thresholding, but they carried out the subsequent differential expression analysis in different ways: Gasperini fitted negative binomial regression models to the data, whereas Xie applied nonparametric tests of independence. We study Gasperini's variant of the thresholding method, as it relates more closely to GLM-EIV.

Let $n \in \mathbb{N}$ be the number of cells assayed in the experiment. Consider a single perturbation and a single gene. For cell $i \in \{1, \dots, n\}$, let $m_i \in \mathbb{N}$ be the number of gene transcripts sequenced; let $g_i \in \mathbb{N}$ be the number of gRNA transcripts sequenced; let $d^m_i \in \mathbb{N}$ be the number of gene transcripts sequenced across \textit{all} genes (i.e., the library size or sequencing depth); and finally, let $z_i \in \mathbb{R}^{d-1}$ be the cell-specific technical factors (e.g., sequencing batch, percent mitochondrial reads, etc.) The letters ``m,'' ``g'', and ``d'' stand for ``mRNA,'' ``gRNA,'' and ``depth,'' respectively. The thresholding method is defined as follows:
\begin{itemize}
	\item[1.] For a given threshold $c \in \mathbb{N}$, let the imputed perturbation assignment $\hat{p}_i \in \{0, 1\}$ be given by $\hat{p}_i = 0$ if  $g_i < c$ and $\hat{p}_i = 1$ otherwise.
	\item[2.] Assume that $m_i$ is related to $\hat{p}_i, d^m_i,$ and $z_i$ through the following GLM:
	\begin{equation}\label{thresh_glm}
	m_i | (\hat{p}_i, z_i, d^m_i) \sim \textrm{NB}_{\theta^m}(\mu_i); \quad
	\log(\mu_i) = \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i + \log\left(d_i^m\right),
	\end{equation}
	where (i) $NB_{\theta^m}(\mu_i)$ is a negative binomial distribution with mean $\mu_i$ and known size parameter $\theta^m$; (ii) $\beta^m_0 \in \mathbb{R}, \beta^m_1 \in \mathbb{R},$ and $\gamma_m \in  \mathbb{R}^{d-1}$ are unknown parameters; and (iii) $\log(d_i^m)$ is an offset term.
